---
title: "Assignment 1"
subtitle: "Dataset Selection and Initial Processing - GSE273848"
author: "Hyunwoo Kwon"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    fig_caption: yes
bibliography: Assignment1.bib
---

# Selecting the Dataset

### **Why is the dataset of interest to you?**
Esophageal adenocarcinoma (EAC) is a highly lethal cancer with poor prognosis. The study "Definition of a multi-omics signature for Esophageal Adenocarcinoma prognosis prediction [totalRNA-seq]" (GSE273848) investigates transcriptional signatures from immune cells to improve prognosis prediction. Understanding the immune response in EAC can help identify biomarkers for patient stratification and potential therapeutic targets.

For this course, I searched the GEO database using the keyword "Esophageal Adenocarcinoma" to find an RNA-seq dataset relevant to immune response and prognosis. I used filters for "Expression profiling by high throughput sequencing" and "Homo sapiens" as the organism type.

I selected GSE273848, which profiles immune cells from EAC patients using single-cell RNA sequencing (scRNA-seq) and bulk RNA-seq. The study identifies transcriptional changes associated with prognosis.

### **Control and Test Conditions**
The dataset contains tumor and matched non-tumor samples from therapy-naïve EAC patients. The control condition consists of non-tumor tissue samples, while the test condition consists of tumor tissue samples. In total, there are 23 samples with transcriptomic profiling.

# Initial Processing

## **Load Required Packages**
```{r, message=FALSE}
# Load necessary packages for data retrieval and processing
BiocManager::install("GEOquery")
BiocManager::install("edgeR")
BiocManager::install("biomaRt")
library("GEOquery") # For querying and downloading from the GEO database
library("edgeR") # For filtering and normalization of RNA-seq data
library("biomaRt") # For mapping Ensembl IDs to HUGO symbols
```

## **Retrieve Dataset Information**
```{r message=FALSE}
# Define the GEO ID for the dataset and retrieve it using GEOquery
dataset_geoID <- "GSE273848"
gse <- getGEO(dataset_geoID, GSEMatrix=FALSE)
```

The dataset summary provides an overview of the experiment, including the conditions and sample metadata:
```{r message = FALSE}
gse@header$summary
```

## **Collecting Sample Annotation Data**
```{r}
# Extract metadata for each sample and format it into a structured dataframe
list_of_samples <- gse@gsms
sample_info <- do.call(rbind, lapply(list_of_samples,function(x){c(x@header$title, x@header$characteristics_ch1)}))
sample_info <- as.data.frame(sample_info)
colnames(sample_info) <- c("header", "patient ID", "sample type", "treatment")
```

### **Filtering for Tumor and Non-Tumor Samples**
```{r}
# Clean sample descriptions for easier filtering
sample_info[,"sample type"] <- gsub("sample type: ", "", sample_info[,"sample type"])
sample_info[,"treatment"] <- gsub("treatment: ", "", sample_info[,"treatment"])

# Filter the dataset to retain only tumor and non-tumor samples
sample_info <- sample_info[sample_info$sample.type %in% c("Tumor", "Non-Tumor"), ]
```

## **Download the Expression Data**
```{r message = FALSE}
library(R.utils) # Needed for gunzip()

data_filename <- getGEOSuppFiles(dataset_geoID, fetch_files = FALSE)$fname[1]
print(paste("Data filename:", data_filename))

if (!file.exists("raw_counts.RData")) {
  print("Downloading GEO supplementary file...")
  sfiles = getGEOSuppFiles(dataset_geoID, filter_regex = data_filename, fetch_files = TRUE)
  
  # Check if the file exists
  downloaded_file <- file.path(dataset_geoID, data_filename)
  print(paste("Downloaded file:", downloaded_file))
  
  # Unzip the file (since it's .gz, not a .tar archive)
  extracted_file <- gsub(".gz$", "", downloaded_file) # Remove .gz extension
  gunzip(downloaded_file, destname = extracted_file, remove = FALSE)

  print("Extracted file:")
  print(extracted_file)

  # Read the unzipped file
  raw_count_data <- read.delim(extracted_file, header=TRUE)
  print("Raw data preview:")
  print(head(raw_count_data))

  rownames(raw_count_data) <- raw_count_data[,1]
  raw_count_data <- raw_count_data[, -1]
  save(raw_count_data, file = "raw_counts.RData")
} else {
  load("raw_counts.RData")
}
```

# **Normalization and Quality Control**

### **Pre-Normalization Statistics**
Before normalization, we examine the distribution of raw counts to assess data quality.
```{r}
summary(raw_count_data)

# Add a pseudocount (+1) to avoid log2(0) issues
boxplot(log2(cpm(raw_count_data) + 1), main="Raw Count Distribution", col="lightblue")
```

### **Filtering Lowly Expressed Genes**
To remove noise, we apply a filtering threshold to exclude genes with low expression across samples.
```{r}
filtered_counts <- raw_count_data[apply(cpm(raw_count_data), 1, function(x){sum(x > 1) >= 4}), ]
```

### **Normalization Using TMM**
TMM (Trimmed Mean of M-values) normalization is used to correct for sequencing depth differences.
```{r}
d <- DGEList(counts=filtered_counts)
d <- calcNormFactors(d, method="TMM")
normalized_counts <- cpm(d, log=TRUE)
```

### **Post-Normalization Analysis**
```{r}
summary(normalized_counts)
boxplot(normalized_counts, main="Normalized Data Distribution", col="lightgreen")
```

# **Mapping Gene Identifiers to HUGO Symbols**
```{r}
# Map Ensembl gene IDs to HUGO symbols
# Load biomaRt
library(biomaRt)

# Connect to Ensembl (use latest dataset)
ensembl_human <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl") 
#following code kept on causing error. 

# Retrieve mapping of Ensembl Gene IDs to HUGO Symbols
id_conversion <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id",
                       values = rownames(normalized_counts),
                       mart = ensembl_human)

# Merge the mapped IDs with the normalized counts data
normalized_counts_annot <- merge(id_conversion, normalized_counts, by.x = 1, by.y = 0, all.y = TRUE)

# Remove unmapped rows
normalized_counts_annot <- normalized_counts_annot[!is.na(normalized_counts_annot$hgnc_symbol), ]
```

# **Final Dataset Summary and Questions Answered**
- **How many samples are in each condition?** The dataset includes 22 tumor and matched 18 non-tumor samples.
- **Handling non-unique gene mappings?** There were couple duplicate genes shown in the samples. For the following sets of the data I have aggregated them by taking their mean expression.
- **Handling unmapped genes?** Unmapped gene identifiers were discarded.
- **Outliers and handling?** As I went through the original paper regarding the following stduy, I was not able to identify any action of removing the outliers. Therefore, I just did not do additional filtering regarding outliers.
- **Handling replicates?** Replicates were retained separately for further analysis.
- **Final dataset coverage?** Approximately 20386 genes remain after filtering and normalization.

---
